INGRID: Creating Languages in MPS from ANTLR Grammars

### 2017-06-14 chapters/background.tex
editing

In this section, first we introduce basic features of the MPS platform, and then we discuss related work, including other existing projects that aim to add support for general purpose languages into MPS \todo{IDE based on projectional editor}.

@@

\todo{For each MPS term (Structure, Editor, TextGen, concept, and so on) say which generally/widely known terms (from related academic work) are similar/analogous. Relate the MPS-specific terms to the large body of academic work.}

\todo{Napsat sem nebo pripadne do Introduction (ale radsi sem): MPS platform, and tools built upon it (e.g., mbeddr) addresses to a large degree known limitations of projectional editors, such as usability of editors for languages based on textual notation (e.g., non-linear typing) and support for version control [citovat "M. Voelter, J. Siegmund, T. Berger, and B. Kolb. Towards User-Friendly Projectional Editors. SLE 2014, LNCS 8706"].}

\todo{Mozna pouzit termin "language concept" misto samotneho "concept".}

JetBrains MPS is a language workbench --- an IDE that allows developers to create their own languages and use them to write code.
The code can then be transformed into a target language, typically a GPL, such as Java or C, and eventually compiled into executable programs.

@@

The building blocks of MPS models are called nodes.
Code of any program in MPS is built from nodes, which represent instances of concepts defined in one of the languages that the program declares to be using.
\todo{zvyraznit slovo "concept"} In the context of MPS, a concept is, in fact, a building block of a language definition.
We use the terms \emph{MPS concept} and \emph{AST node} when needed to avoid confusion.

@@

The definition of an MPS concept (language element) consists of several aspects, where each aspect codifies a different part of the AST nodes' behavior.
The essential aspects are the following: \emph{Structure} (abstract syntax), \emph{Editor} (concrete syntax) and \emph{TextGen}.
Editor defines the concrete syntax (i.e., how the code is visualized and edited) and TextGen specifies how AST nodes are transformed into textual representation.
If, instead of generating text directly, programs in the language are supposed to be transformed into another language that is available in MPS, the Generator aspect must be used to specify the model-to-model transformation rules.
Since only the Structure, Editor and TextGen aspects are relevant for the contribution of this paper, we describe these three below in more detail, and neglect other aspects such as type system and data-flow.

@@

\paragraph{TextGen.}
The TextGen aspect specifies how a given AST node will be translated into plain text representation.
It is typically needed only for the bottom-line base languages.
DSLs, on the other hand, need to define rules for model-to-model conversions (Generators), since these are rarely converted to text directly.
After TextGen has generated textual sources from an AST, a compiler for the particular GPL is invoked to compile the textual sources into binary code.
The TextGen definition follows a very straightforward pattern --- each node outputs its textual representation into a buffer, while calling TextGen of its children nodes at the right moments.
MPS calls the corresponding method on the root AST nodes of the given program.

@@

\subsection{Related Projects}
\label{sect:RELATED}

The BaseLanguage~\cite{ref:BASELANG} is an almost full port of Java, which was created manually by JetBrains and extended with MPS-specific features.
% and it is still undergoing changes as Java itself is evolving.
The C language has also been manually tailored for MPS within the mbeddr project~\cite{ref:MBEDDR}.

Besides that, we are aware of three other projects that provide certain support for GPLs in the context of MPS --- namely PE4MPS, ANTLR{\_}MPS, and mps-metabnf.
We describe their main features and limitations in this section.

@@

@@

\todo{
S. Erdweg, T. Rendel, C. K\"{a}stner, and K. Ostermann. Layout-Sensitive Generalized Parsing. In Proceedings of SLE 2012, LNCS 7745.
	Source code layout specified declaratively through constraints on shape and relative positions of syntax trees (AST).
	Layout constraints are represented/encoded using annotations on production rules of context-free grammars.
	Useful for layout-sensitive programming language (e.g., Python).
	Their proposed approach: grammar rule applies only if the input satisfied the constraint (on shape and positioning of source code fragments).
		Parsing algorithm that accepts only inputs with valid layout (that satisfies the constraints attached to grammar rules).
	Constraints support (can express): indentation, number of line breaks, alignment.
	Hand-coded approach used in Python (how the annotated grammar is processed) --- mentioned for comparison.
		Special tokens that indicate layout aspects (indentation, alignment) are added to stream of tokens by preprocessor.
		Layout handled by algorithm that works with syntax trees (ASTs) and token streams.
	The main difference from our approach (INGRID): layout inferred automatically (via automated learning) from several input files.
		No layout annotations in the grammar are used.
		Currently, simple heuristics are used to create editor and textgen.
}

\todo{
Mention other state-of-the-art language workbenches: Spoofax.
	Reference: L.C.L. Kats and E. Visser. The Spoofax Language Workbench: Rules for Declarative Specification of Languages and IDEs. OOPSLA 2010
	Eclipse-based IDE support for SDF-based languages.
	Parser-based language workbench.
		Syntax essentially restricted to linear text.
		But it supports language extension and composition.
}

>>>>

In this section, first we introduce basic features of the MPS platform, and then we discuss related work, including other existing projects that aim to add support for general purpose languages into MPS.

@@

JetBrains MPS is a language workbench --- an IDE that allows developers to create their own languages and use them to write code.
The code can then be transformed into a target language, typically a GPPL, such as Java or C, and eventually compiled into executable programs.

@@

The building blocks of MPS models are called nodes.
Code of any program in MPS is built from nodes, which represent instances of concepts from the languages that the program declares to be using.
In the context of MPS, a \emph{concept} is a language element, i.e. a building block of a language definition.
We use the terms \emph{MPS concept} and \emph{AST node} when needed to avoid confusion.

@@

The definition of an MPS concept (language element) consists of several aspects, where each aspect codifies a different part of the AST nodes' behavior.
The essential aspects are the following: \emph{Structure}, \emph{Editor} and \emph{TextGen}.
Structure represents the abstract syntax (types and hierarchy of AST nodes), Editor defines the concrete syntax (i.e., how the code is visualized and edited) and TextGen specifies how AST nodes are transformed into textual representation.
If, instead of generating text directly, programs in the language are supposed to be transformed into another language that is available in MPS, the Generator aspect must be used to specify the model-to-model transformation rules.
Since only the Structure, Editor and TextGen aspects are relevant for the contribution of this paper, we describe these three below in more detail, and neglect other aspects such as type system and data-flow.

@@

\paragraph{TextGen.}
The TextGen aspect specifies how a given AST node will be translated into plain text representation.
It is typically needed only for the bottom-line base languages.
DSLs, on the other hand, need to define rules for model-to-model conversions (Generators), since programs in such languages are rarely converted to text directly.
After TextGen has generated textual sources from an AST, a compiler for the particular GPPL is invoked to compile the textual sources into binary code.
The TextGen definition follows a very straightforward pattern --- each node outputs its textual representation into a buffer, while calling TextGen of its children nodes at the right moments.
MPS calls the corresponding method on the root AST nodes of the given program.

@@

\subsection{Related Work}
\label{sect:RELATED}

We are aware just of a few attempts to create ports of general-purpose languages into IDE tools based on projectional editing.
All the ports that we mention below were conducted manually.

The BaseLanguage~\cite{ref:BASELANG} is an almost full port of Java, which was created by JetBrains and extended with MPS-specific features.
% and it is still undergoing changes as Java itself is evolving.
The C language has also been manually tailored for MPS within the mbeddr project~\cite{ref:MBEDDR}.

The following three projects --- PE4MPS, ANTLR{\_}MPS, and mps-metabnf --- aim to provide certain support for GPPLs within the MPS platform.
We describe their main features and limitations in this section.

@@

We discuss application of PE4MPS on JavaScript in Section~\ref{sect:EVAL}, and compare its results to our \textsc{Ingrid} approach.

@@

\paragraph{Other IDEs.}
Note that while the \textsc{Ingrid} method has been designed primarily for MPS, and can be easily adapted to other IDE tools based on projectional editing, there also exist several language workbenches that use parser generators (such as ANTLR) and support free editing of textual source code.
A prominent member of the latter category of tools is Spoofax~\cite{ref:KV10}, which essentially restricts the syntax of DSLs to linear text.
\textsc{Ingrid} is not applicable for creating languages inside such tools.

### 2017-06-14 chapters/language_definition.tex
editing

The proposed \textsc{Ingrid} method accepts grammars in the ANTLR v4 notation~\cite{ref:ANTLRBOOK} as input.
\todo{Zmenit nasledujici vetu: There exist ANTLR v4 grammars for all widely used programming languages ...}
All widely used programming languages have their syntax defined using this notation~\footnote{https://github.com/antlr/grammars-v4}.
Unlike some of the related projects, we did not extend the notation with any custom features, and we also did not create an MPS language for the ANTLR notation.

@@

Our approach differs from the related projects (Section~\ref{sect:RELATED}) especially in the level of automation. It also has much better support for the definition of Editor and TextGen.

@@

Every tree that represents a lexer rule is flattened into the equivalent regular expression by the recursive algorithm in Figure~\ref{fig:ALGFLATTEN}.
A sequence is created from the elements of each alternative, and then all those sequences are joined by the alternation operator \code{$|$} to form a regular expression.
\todo{Algorithm seems to fail if the grammar has right-recursion (according to one reviewer) --- check that and describe how we handle that case.}

@@

When designing the process for translating AST nodes (i.e., grammar rules) into the MPS language structure, we faced several challenges.
The main challenge related to the structure, as we already mentioned in the introduction, is that a grammar typically contains rules that do not directly correspond to programming language constructs.
Such rules exist at the intermediate layers of a syntax hierarchy.
Their main purpose is to enable easier understanding and maintenance of the grammar by humans.
However, presence of the intermediate layers significantly complicates usage of the given language in MPS, and the layers are actually not necessary for construction of an MPS language.
We show examples illustrating this problem, which we call a \emph{layer problem}, later in the subsection where we describe our solution \todo{Pridat odkaz na konkretni section a opravit gramatickou chybu}.

@@

The key idea of this approach is to skip all the intermediary layers (nodes) in the syntax tree and consider just the leaf nodes, e.g., to be directly offered to the user through the auto-completion menu.
Specifically, the \antlrparserrule{content} rule from the SimpleXML grammar, which is given in Figure~\ref{fig:CONTENTHIERARCHY} together with rules that determine the relevant fragment of the syntax hierarchy, expands ultimately into the leaf nodes that are highlighted using the bold font in Figure~\ref{fig:CONTENTEXPAND}.

@@

For the input that consists of a particular AST node $N$ and the grammar rule $R$ that expands $N$, the procedure implementing the shortcut approach systematically traverses the parser tree (AST) built in the earlier phases in order to identify each AST node that may appear at the end of some derivation chain starting by the rule $R$ from the node $N$.
Such nodes cannot expand further based on any rule in the grammar, and for that reason we call them \emph{end nodes}. \todo{Srovnat pojem "end nodes" vzhledem ke terminal symbols from grammars (jako analogie). Mozna existuje dalsi vhodny termin.}

@@

We experimented with several heuristics to derive a useful code layout, but all of them produced rather suboptimal results \todo{Give examples or quantify this result}.
However, we observed that the most tedious and error-prone step in the manual definition of the Editor aspect is the creation of cells for all literals (keywords), properties, children, and other fields of a given concept that should appear in its visual representation.
This step can be very easily automated.

Our solution that we implemented in the current version of the \textsc{Ingrid} method is to create all the cells and place them in a single row.
The resulting basic layout is illustrated on the example of the \mpsconcept{Element{\_}1} concept in the upper left corner of Figure~\ref{fig:EDITORADJUST}.
Further adjustments of the layout, such as indentation and line breaks, can be done very efficiently in the MPS IDE.
The bottom right part of Figure~\ref{fig:EDITORADJUST} shows a fully customized layout for the \mpsconcept{Element{\_}1} concept.

@@

Based on our experience, it takes a very short time to manually adjust the layout into a form much better than any fully automated heuristic could achieve.
We expect that a typical user of the \textsc{Ingrid} method will be able to use the projectional editor in MPS quite efficiently.
More details are provided in Section~\ref{sect:EVAL}, including our experience with several mainstream programming languages.

@@

\subsection{Remarks about Grammars}

>>>>

The proposed \textsc{Ingrid} method accepts grammars in the ANTLR v4 notation~\cite{ref:ANTLRBOOK} as input.
Syntax definition in the form of an ANTLR v4 grammar exist for all widely used programming languages~\footnote{https://github.com/antlr/grammars-v4}.
Unlike some of the related projects, we did not extend the notation with any custom features, and we also did not create an MPS language for the ANTLR notation.

@@

Our approach differs from the existing projects with similar goals (Section~\ref{sect:RELATED}) especially in the level of automation. It also has much better support for the definition of Editor and TextGen.

@@

Every tree that represents a lexer rule is flattened into the equivalent regular expression by the recursive algorithm in Figure~\ref{fig:ALGFLATTEN}.
A sequence is created from the elements of each alternative, and then all those sequences are joined by the alternation operator \code{$|$} to form a regular expression.

@@

When designing the process for translating AST nodes (i.e., grammar rules) into the MPS language structure, we faced several challenges.
The main challenge related to the structure, as we already mentioned in the introduction, is that a grammar typically contains rules that do not directly correspond to programming language constructs.
Such rules exist at the intermediate layers of a syntax hierarchy.
Their main purpose is to enable easier understanding and maintenance of the grammar by humans.
However, presence of the intermediate layers significantly complicates usage of the given language in MPS, and the layers are actually not necessary for construction of an MPS language.
We show examples illustrating this problem, which we call a \emph{layer problem}, later in this section at a point where we describe our solution.

@@

The key idea of this approach is to skip all the intermediary layers (nodes) in the syntax tree, and consider just the nodes to be directly offered to the user through the auto-completion menu.
Specifically, the \antlrparserrule{content} rule from the SimpleXML grammar, which is given in Figure~\ref{fig:CONTENTHIERARCHY} together with rules that determine the relevant fragment of the syntax hierarchy, expands ultimately into the nodes that are highlighted using the bold font in Figure~\ref{fig:CONTENTEXPAND}.

@@

For the input that consists of a particular AST node $N$ and the grammar rule $R$ that expands $N$, the procedure implementing the shortcut approach systematically traverses the parser tree (AST) built in the earlier phases in order to identify each AST node that (i) may appear in some derivation chain starting by the rule $R$ from the node $N$ and (ii) does not belong to any intermediary layer.
We call them \emph{end nodes}.
Here we must emphasize that end nodes do not correspond to terminal symbols from grammars, because they may represent AST nodes with children (i.e., syntax elements that can be further expanded using grammar rules).

@@

We observed that the most tedious and error-prone step in the manual definition of the Editor aspect is the creation of cells for all literals (keywords), properties, children, and other fields of a given concept that should appear in its visual representation.
This step can be very easily automated.

Our solution that we implemented in the current version of the \textsc{Ingrid} method is to create all the cells and place them in a single row.
The resulting basic layout is illustrated on the example of the \mpsconcept{Element{\_}1} concept in the upper left corner of Figure~\ref{fig:EDITORADJUST}.
Further adjustments of the layout, such as indentation and line breaks, can be done very efficiently by the user in the MPS IDE.
The bottom right part of Figure~\ref{fig:EDITORADJUST} shows a fully customized layout for the \mpsconcept{Element{\_}1} concept.

@@

Based on our experience, it takes a very short time to manually adjust the layout into a form much better than any fully automated heuristic could achieve.
A typical user of the \textsc{Ingrid} method will be able to use the projectional editor in MPS quite efficiently.
We actually experimented with several heuristics to derive a useful code layout, but all of them produced rather suboptimal results --- it is quite hard to automatically identify language concepts for which the default layout should be refactored, and the user would have to adjust the layout by hand anyway.
More details are provided in Section~\ref{sect:EVAL}, including our experience with several mainstream programming languages.

@@

\subsection{Remarks about Grammars}
\label{sect:REMARKSGRAMMARS}
